Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-06-27 Manuscript/01-Intro.Rmd
Revise method section

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we impose a new bi-part spike-and-slab spline prior on the smooth functions to perform bilevel selection, where the linear and nonlinear spaces of smooth functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. Additionally, two fast computing optimization based algorithms are developed for high and ultra-high dimension settings respectively. Meanwhile, to avoid the over-penalization of the basis function coefficients, smoothness penalties are incorporated in the model via re-parameterization. Smoothness penalties are commonly implemented in GAMs through smoothing and penalized regression splines. They are quadratic norm of the coefficients, and allow locally adaptive penalties on each smoothing function. For the most common case of cubic splines, the smoothness penalty conditioning on a smoothness penalty parameter $\lambda_j$ is a function of the integration of second derivative of the spline function, expressed mathematically as
$$
  \text{pen}f_j(x) = \lambda_j \int f^{"}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j,
$$
where $\bs S_j$ is a known penalty matrix, and $\bs \beta_j$ are the basis function coefficients. Smoothness penalties are also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothness penalty allows the separation of the linear space of the smooth function, also know as the null space, from the nonlinear space. @Marra2011 introduced a re-parameterization of the basis function matrix based on the decomposition of the smoothness penalty matrix $\bs S$. The re-parameterization allows different penalties imposed on the null and nonlinear spaces separately, and further allows the sparsity in linear space and smoothness in nonlinear effect interpolation. [TODO: a sentence that it decompose the prior to independent priors, provide obvious computational convenience.] The computational convenience improved the scalability of the model.

>>>>

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we impose a new bi-part spike-and-slab spline prior on the smooth functions to perform bilevel selection, where the linear and nonlinear spaces of smooth functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. Additionally, two fast computing optimization based algorithms are developed for high and ultra-high dimension settings respectively. Meanwhile, to avoid the over-penalization of the basis function coefficients, smoothness penalties are incorporated in the model via re-parameterization. Smoothness penalties are commonly implemented in GAMs through smoothing and penalized regression splines. They are quadratic norm of the coefficients, and allow locally adaptive penalties on each smoothing function. For the most common case of cubic splines, the smoothness penalty conditioning on a smoothness penalty parameter $\lambda_j$ is a function of the integration of second derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{"}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j ,
\end{equation}
where $\bs S_j$ is a known penalty matrix, and $\bs \beta_j$ are the basis function coefficients. Smoothness penalties are also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothness penalty allows the separation of the linear space of the smooth function, also know as the null space, from the nonlinear space. @Marra2011 introduced a re-parameterization of the basis function matrix based on the decomposition of the smoothness penalty matrix $\bs S$. The re-parameterization allows different penalties imposed on the null and nonlinear spaces separately, and further allows the sparsity in linear space and smoothness in nonlinear effect interpolation. [TODO: a sentence that it decompose the prior to independent priors, provide obvious computational convenience.] The computational convenience improved the scalability of the model.

### 2021-06-27 Manuscript/02-Method.Rmd
Revise method section

# Spike-and-slab Additive Model

Following the notation in the previous section \ref{Notation}, suppose we have the model matrix $\bs X$ derived from spline functions of choice $h_j, j = 1, \dots, p.$ for the $p$ covariates that are modeled nonparametrically, the total dimension of $\bs X$ is $n \times \sum\limits_{j=1}^p k_j$. Column-wise centering and re-parameterization introduced in the GAM as random effect section[^5] are performed. The re-parameterization has two purposes. First of all the re-parameterization reshapes the columns to be linearly independent to each other, and hence, independent prior can be set to the coefficients. Secondly, the re-parameterization decomposes the model matrix $\bs X_j$ for the $j$th covariates to two sub-matrices representing the null space of the spline and penalized space of the spline,
$$
\bs X_j = \begin{bmatrix}
\bs X_j^{0} & \bs X_j^{pen}
\end{bmatrix},
$$
where $\bs X_j^{null}$ is the column space representing the null space of the spline model, and $\bs X_j^{pen}$ denotes the column spaces that need to be shrunk for smoothness of the spline. The purpose for this re-parameterization is to isolate the null space from the smoothing space, such that answering the question, if non-linear relationship is necessary, becomes easy by deciding if the corresponding coefficients of the $\bs X_j^{pen}$ are jointly zero. This strategy is widely used, previously mentioned in @Scheipl2012 and the references there in.^[Possibly select some to add here.] Hereafter, we denote the coefficients of the null space as $\bs \beta^{0}_j$ and the coefficients of the penalized space as $\bs \beta^{pen}_j$. The spline function for the $j$the covariate is 
$$
f_j(X) = f_j^0(X) + f_j^{pen}(X) = {\beta_j^0}^T X_j^0 + {\beta_j^{pen}}^T X_j^{pen},
$$
where the superscript $^t$ denotes matrix transpose.

Without loss of generality, we assume that the model doesn't include the the parametric part of the model $\bs A$. The coefficients of $\bs A$ can be estimated easily in the proposed EM algorithms by local update conditioning on the temporary value of the spline coefficients.

## Spike and Slab Prior

>>>>

# Bayesian Hierarchical Additive Models

Following the GLM notation introduced in Section \label{introduction}, we have a generalized additive model 
$$
E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
$$
with the smoothing function $f_j(x_j)$ of the variable $x_j, j = 1, \dots, p.$ The basis function matrix, i.e. the design matrix derived from the smoothing function $f_j(x_j)$, is denoted $\bs X_j$ for the variable $x_j$. The dimension of the design matrix depends on the choice of the smoothing function. For example, a cubic spline function with $k$ knots produces a $k+2$ elements column vector. Its corresponding smoothness penalty matrix is denoted as $\bs S_j$. In all, the GAMs can be formulated in a matrix form 
$$
E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j),
$$
where $\bs \beta_j$ denotes the basis function coefficients for the $j$th variable, and the matrix transposing operation is denoted with a superscript $^t$. To note, the proposed model can include parametric forms of variables in the model, treating the parametric function as a special case of the smoothing function, e.g. $f_j(x_j) = x_j$ with the smoothness penalty matrix defined as $\bs S_j = \begin{bmatrix}0\end{bmatrix}$.

To encourage proper smoothness of the smoothing functions, we adopt the idea of the smoothing penalty from smoothing spline models. The basic idea is to set up a smoothing penalty, depending on the smoothness penalty matrix via Equation \ref{eq:smoothpen}, in the posterior density function, or least squared and likelihood function in the frequentist framework. However, direct integration of the smoothness penalty into the sparsity penalty is not obvious. Given that the smoothness penalty matrix is, @Marra2011 introduced a re-parameterization of the basis function matrix $\bs X$ based on the decomposition of the smoothness penalty matrix $\bs S$.

 Column-wise centering and re-parameterization introduced in the GAM as random effect section[^5] are performed. The re-parameterization has two purposes. First of all the re-parameterization reshapes the columns to be linearly independent to each other, and hence, independent prior can be set to the coefficients. Secondly, the re-parameterization decomposes the model matrix $\bs X_j$ for the $j$th covariates to two sub-matrices representing the null space of the spline and penalized space of the spline,
$$
\bs X_j = \begin{bmatrix}
\bs X_j^{0} & \bs X_j^{pen}
\end{bmatrix},
$$
where $\bs X_j^{null}$ is the column space representing the null space of the spline model, and $\bs X_j^{pen}$ denotes the column spaces that need to be shrunk for smoothness of the spline. The purpose for this re-parameterization is to isolate the null space from the smoothing space, such that answering the question, if non-linear relationship is necessary, becomes easy by deciding if the corresponding coefficients of the $\bs X_j^{pen}$ are jointly zero. This strategy is widely used, previously mentioned in @Scheipl2012 and the references there in.^[Possibly select some to add here.] Hereafter, we denote the coefficients of the null space as $\bs \beta^{0}_j$ and the coefficients of the penalized space as $\bs \beta^{pen}_j$. The spline function for the $j$the covariate is 
$$
f_j(X) = f_j^0(X) + f_j^{pen}(X) = {\beta_j^0}^T X_j^0 + {\beta_j^{pen}}^T X_j^{pen}.
$$

## Spike and Slab Spline Prior

### 2021-06-27 Manuscript/SS_GAM.tex
Revise method section

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we impose a new bi-part spike-and-slab spline prior on the
smooth functions to perform bilevel selection, where the linear and
nonlinear spaces of smooth functions can be selected separately. The
prior setup encourages a flexible solution, rending one of three
possibilities: no effect, only linear effect, or nonlinear effect of a
predictor. Additionally, two fast computing optimization based
algorithms are developed for high and ultra-high dimension settings
respectively. Meanwhile, to avoid the over-penalization of the basis
function coefficients, smoothness penalties are incorporated in the
model via re-parameterization. Smoothness penalties are commonly
implemented in GAMs through smoothing and penalized regression splines.
They are quadratic norm of the coefficients, and allow locally adaptive
penalties on each smoothing function. For the most common case of cubic
splines, the smoothness penalty conditioning on a smoothness penalty
parameter \(\lambda_j\) is a function of the integration of second
derivative of the spline function, expressed mathematically as \[
  \text{pen}f_j(x) = \lambda_j \int f^{"}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j,
\] where \(\boldsymbol{S}_j\) is a known penalty matrix, and
\(\boldsymbol{\beta}_j\) are the basis function coefficients. Smoothness
penalties are also previously used in the spike-and-slab GAM
\citep{Scheipl2012} and the sparsity-smoothness penalty
\citep{Meier2009}. Moreover, incorporating the smoothness penalty allows
the separation of the linear space of the smooth function, also know as
the null space, from the nonlinear space. \citet{Marra2011} introduced a
re-parameterization of the basis function matrix based on the
decomposition of the smoothness penalty matrix \(\boldsymbol{S}\). The
re-parameterization allows different penalties imposed on the null and
nonlinear spaces separately, and further allows the sparsity in linear
space and smoothness in nonlinear effect interpolation. {[}TODO: a
sentence that it decompose the prior to independent priors, provide
obvious computational convenience.{]} The computational convenience
improved the scalability of the model.

@@

\hypertarget{spike-and-slab-additive-model}{%
\section{Spike-and-slab Additive
Model}\label{spike-and-slab-additive-model}}

Following the notation in the previous section \ref{Notation}, suppose
we have the model matrix \(\boldsymbol{X}\) derived from spline
functions of choice \(h_j, j = 1, \dots, p.\) for the \(p\) covariates
that are modeled nonparametrically, the total dimension of
\(\boldsymbol{X}\) is \(n \times \sum\limits_{j=1}^p k_j\). Column-wise
centering and re-parameterization introduced in the GAM as random effect
section{[}\^{}5{]} are performed. The re-parameterization has two
purposes. First of all the re-parameterization reshapes the columns to
be linearly independent to each other, and hence, independent prior can
be set to the coefficients. Secondly, the re-parameterization decomposes
the model matrix \(\boldsymbol{X}_j\) for the \(j\)th covariates to two
sub-matrices representing the null space of the spline and penalized
space of the spline, \[
\boldsymbol{X}_j = \begin{bmatrix}
\boldsymbol{X}_j^{0} & \boldsymbol{X}_j^{pen}
\end{bmatrix},
\] where \(\boldsymbol{X}_j^{null}\) is the column space representing
the null space of the spline model, and \(\boldsymbol{X}_j^{pen}\)
denotes the column spaces that need to be shrunk for smoothness of the
spline. The purpose for this re-parameterization is to isolate the null
space from the smoothing space, such that answering the question, if
non-linear relationship is necessary, becomes easy by deciding if the
corresponding coefficients of the \(\boldsymbol{X}_j^{pen}\) are jointly
zero. This strategy is widely used, previously mentioned in
\citet{Scheipl2012} and the references there in.\footnote{Possibly
  select some to add here.} Hereafter, we denote the coefficients of the
null space as \(\boldsymbol{\beta}^{0}_j\) and the coefficients of the
penalized space as \(\boldsymbol{\beta}^{pen}_j\). The spline function
for the \(j\)the covariate is \[
f_j(X) = f_j^0(X) + f_j^{pen}(X) = {\beta_j^0}^T X_j^0 + {\beta_j^{pen}}^T X_j^{pen},
\] where the superscript \(^t\) denotes matrix transpose.

@@

\hypertarget{spike-and-slab-prior}{%
\subsection{Spike and Slab Prior}\label{spike-and-slab-prior}}

@@

\hypertarget{spike-and-slab-spline-prior}{%
\subsection{Spike and slab spline
prior}\label{spike-and-slab-spline-prior}}

>>>>

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we impose a new bi-part spike-and-slab spline prior on the
smooth functions to perform bilevel selection, where the linear and
nonlinear spaces of smooth functions can be selected separately. The
prior setup encourages a flexible solution, rending one of three
possibilities: no effect, only linear effect, or nonlinear effect of a
predictor. Additionally, two fast computing optimization based
algorithms are developed for high and ultra-high dimension settings
respectively. Meanwhile, to avoid the over-penalization of the basis
function coefficients, smoothness penalties are incorporated in the
model via re-parameterization. Smoothness penalties are commonly
implemented in GAMs through smoothing and penalized regression splines.
They are quadratic norm of the coefficients, and allow locally adaptive
penalties on each smoothing function. For the most common case of cubic
splines, the smoothness penalty conditioning on a smoothness penalty
parameter \(\lambda_j\) is a function of the integration of second
derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{"}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known penalty matrix, and
\(\boldsymbol{\beta}_j\) are the basis function coefficients. Smoothness
penalties are also previously used in the spike-and-slab GAM
\citep{Scheipl2012} and the sparsity-smoothness penalty
\citep{Meier2009}. Moreover, incorporating the smoothness penalty allows
the separation of the linear space of the smooth function, also know as
the null space, from the nonlinear space. \citet{Marra2011} introduced a
re-parameterization of the basis function matrix based on the
decomposition of the smoothness penalty matrix \(\boldsymbol{S}\). The
re-parameterization allows different penalties imposed on the null and
nonlinear spaces separately, and further allows the sparsity in linear
space and smoothness in nonlinear effect interpolation. {[}TODO: a
sentence that it decompose the prior to independent priors, provide
obvious computational convenience.{]} The computational convenience
improved the scalability of the model.

@@

\hypertarget{bayesian-hierarchical-additive-models}{%
\section{Bayesian Hierarchical Additive
Models}\label{bayesian-hierarchical-additive-models}}

Following the GLM notation introduced in Section \label{introduction},
we have a generalized additive model \[
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
\] with the smoothing function \(f_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(f_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+2\)
elements column vector. Its corresponding smoothness penalty matrix is
denoted as \(\boldsymbol{S}_j\). In all, the GAMs can be formulated in a
matrix form \[
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\] where \(\boldsymbol{\beta}_j\) denotes the basis function
coefficients for the \(j\)th variable, and the matrix transposing
operation is denoted with a superscript \(^t\). To note, the proposed
model can include parametric forms of variables in the model, treating
the parametric function as a special case of the smoothing function,
e.g.~\(f_j(x_j) = x_j\) with the smoothness penalty matrix defined as
\(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothness of the smoothing functions, we adopt the
idea of the smoothing penalty from smoothing spline models. The basic
idea is that set up a smoothing penalty, depending on the smoothness
penalty matrix via the penalty introduced in Eq \ref{eq:smoothpen} , in
the posterior density function, or least squared and likelihood function
in the frequentist framework. However, direct integration of the
smoothness penalty with the sparsity penalty is not straightforward.

have the model matrix \(\boldsymbol{X}\) derived from spline functions
of choice \(h_j, j = 1, \dots, p.\) for the \(p\) covariates that are
modeled nonparametrically, the total dimension of \(\boldsymbol{X}\) is
\(n \times \sum\limits_{j=1}^p k_j\). Column-wise centering and
re-parameterization introduced in the GAM as random effect
section{[}\^{}5{]} are performed. The re-parameterization has two
purposes. First of all the re-parameterization reshapes the columns to
be linearly independent to each other, and hence, independent prior can
be set to the coefficients. Secondly, the re-parameterization decomposes
the model matrix \(\boldsymbol{X}_j\) for the \(j\)th covariates to two
sub-matrices representing the null space of the spline and penalized
space of the spline, \[
\boldsymbol{X}_j = \begin{bmatrix}
\boldsymbol{X}_j^{0} & \boldsymbol{X}_j^{pen}
\end{bmatrix},
\] where \(\boldsymbol{X}_j^{null}\) is the column space representing
the null space of the spline model, and \(\boldsymbol{X}_j^{pen}\)
denotes the column spaces that need to be shrunk for smoothness of the
spline. The purpose for this re-parameterization is to isolate the null
space from the smoothing space, such that answering the question, if
non-linear relationship is necessary, becomes easy by deciding if the
corresponding coefficients of the \(\boldsymbol{X}_j^{pen}\) are jointly
zero. This strategy is widely used, previously mentioned in
\citet{Scheipl2012} and the references there in.\footnote{Possibly
  select some to add here.} Hereafter, we denote the coefficients of the
null space as \(\boldsymbol{\beta}^{0}_j\) and the coefficients of the
penalized space as \(\boldsymbol{\beta}^{pen}_j\). The spline function
for the \(j\)the covariate is \[
f_j(X) = f_j^0(X) + f_j^{pen}(X) = {\beta_j^0}^T X_j^0 + {\beta_j^{pen}}^T X_j^{pen},
\] where the superscript \(^t\) denotes matrix transpose.

@@

\hypertarget{spike-and-slab-spline-prior}{%
\subsection{Spike and Slab Spline
Prior}\label{spike-and-slab-spline-prior}}

@@

\hypertarget{spike-and-slab-spline-prior-1}{%
\subsection{Spike and slab spline
prior}\label{spike-and-slab-spline-prior-1}}
